---
title: "IsoformSwitchAnalyzeR_project"
output: html_document
date: "2026-05-02"
---


#Step1: Generating mock dataset
Generating a small,synthesic IsoformSwitchAnalyzeR-like dataset
columns= genes, isoform, condition, expression, IF
```{r}
#structure of the dataframe
genes <- c("TP53","BRCA1", "EGFR") #fixed three genes

isoforms <- list(
  TP53 = c("TP53_1","TP53_2","TP53_3"),
  BRCA1 = c("BRCA1_1", "BRCA1_2"),
  EGFR = c("EGFR_1","EGFR_2","EGFR_3") 
)

conditions <- c("control", "disease") #two conditions; control and disease




#expression data
set.seed(123)
df <-data.frame()

for (g in genes){
  
  for (cond in conditions){
    iso_list <- isoforms[[g]] #list of isoforms in every gene
    n_iso <- length(iso_list) #number of isoforms in each gene
    
    
  #base expression pattern
    if (g == "TP53") {
      expr <- if (cond == "control") c(100, 20, 10) else c(30, 80, 20)
    }
    
    if (g == "BRCA1") {
      expr <- if (cond == "control") c(80, 40) else c(20, 100)
    }
    
    if (g == "EGFR") {
      expr <- if (cond == "control") c(50, 50, 10) else c(10, 90, 20)
    }
    temp <- data.frame(
      gene = g,
      isoform = iso_list,
      condition = cond,
      expression = expr
    )
    
    df <- rbind(df, temp)
  }
}
df
```


##Computing isoform fractions for each entry in the mock dataset
```{r}
library(dplyr)

df <- df %>%
  group_by(gene, condition) %>%
  mutate(IF = expression / sum(expression)) %>%
  ungroup()
#groups the isoform entries by gene and condition and calculates the IF for each gene-condition pair

#saving
write.csv(df, "mock_isoform_switch_data.csv", row.names = FALSE)
df
```
TP53- iso1 was dominant in control cases, switched to iso2 being dominant in disease cases
BRCA1 - iso1 -> iso2 switch
EGFR- both iso1 and 2 were dominant in control-> iso2 dominant in disease cases


#Step2 : Isoform Switching plot
A plot that would show how isoforms for a specific gene switch between control and disease conditions

```{r}
#strong isoform switch in TP53 so filtering out this gene

library(dplyr)
gene_data <- df %>% filter(gene=="TP53")

#making isoform fraction bar plot, this will show the isoform fractions for each isoform in each condition for this gene
library(ggplot2)
ggplot(gene_data, aes(x=condition, y=IF, fill= isoform)) + 
  geom_bar(stat="identity")+
  labs(title= "Isoform usage for TP53",
       x="Condition",
       y="Isoform Fraction")+
  theme_minimal()
```
The plot clearly shows, in control group iso1 dominates while in disease group iso2 dominates. Therefore, isoform switch has occurred.


## Slope plot for isoform switch
```{r}
ggplot(gene_data, aes(x = condition, y = IF, group = isoform, color = isoform)) +
  geom_line(size=1.2)+
  geom_point(size=3)+
  labs(title="Isoform Switching (TP53)" ,
       x= "Isoform Fraction")+
  theme_minimal()

```

## making a function for these slopes
```{r}
plot_switch <- function(data, gene_name) {
  
  gene_data <- data %>% filter(gene == gene_name)
  
  ggplot(gene_data, aes(x = condition, y = IF, group = isoform, color = isoform)) +
    geom_line(size = 1.2) +
    geom_point(size = 3) +
    labs(title = paste("Isoform Switch:", gene_name),
         y = "Isoform Fraction") +
    theme_minimal()
}

#calling the function
plot_switch(df, "BRCA1" )
```




```{r}
Sys.which("git")
```













